% !TEX encoding = UTF-8 Unicode
% !TEX program = lualatex
% was lualatexmk
% don't forget Edit > Substitutions > Text Replacement
% needs file combining_preprocessor.lua, & BibTex.bib (e.g.) for \bibliography
%\input{/Users/wdlinch3/Dropbox/Rashoumon/LaTeX/macros/headerWDL3}
\input{headerWDL3}

\def\red#1{{\color{red}#1}}
\def\blue#1{{\color{blue}#1}}
\def\green#1{{\color{green}#1}}

\usepackage{placeins}

\def\gap{38mm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%  Theorem environments  %%%%%%%%%%%%%%%%%%%%
% Theorem environments
\usepackage{amsthm} %Enable theorem environments
%% No-number theorems
%\theoremstyle{definition}
%\newtheorem*{definition}{Definition}
%\theoremstyle{plain}
%\newtheorem*{prop}{Proposition}
%\newtheorem*{lemma}{Lemma}
%\newtheorem*{thm}{Theorem}
%\newtheorem*{cor}{Corollary}
%\newtheorem*{example}{Example}
%% Numbered theorems
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{conjecture}{Conjecture}
\newtheorem{example}{Example}
\newtheorem{result}{Result}
\newtheorem{claim}{Claim}
\newtheorem*{result*}{Result}
\newtheorem{exercise}{Exercise}
\newtheorem{problem}{Problem}


\begin{document}

{\center
{\color{violet}\fontsize{28pt}{34pt}\bf\sffamily 
Supplemental Material 1: Orthogonal Functions
\\[.5in]
}

%{\color[rgb]{0,.7,1}\today}\\[.5in]}
}
{\abstract
These are notes for a discussion about linear algebra and analysis for the students of the Research Club.
}

%\vfill

%\thispagestyle{empty}
%\newpage

%{
%%\linespread{1.1}
%%\linespread{1}
%%\linespread{.9}
%\linespread{.8}
%\small
%%\tableofcontents
%}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

§{Linear algebra}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Let $V$ be an $𝐑$-vector space, and recall that a linear operator $L$ on $V$ is a linear map
$$
L:V ⟶ V
$$
An eigenvector of $L$ is a vector $v_λ ∈ V$ such that
$$
L v_λ = λ v_λ 
$$
The number $λ ∈ V$ is called the eigenvalue of $L$ associated to the eigenvector $v_λ$. 

Suppose $V$ is equipped with an inner product 
$$
\left< ⋅ ,  ⋅ \right> : V×V ⟶  𝐑 
$$
The adjoint $L^\dagger$ of $L$ with respect to the inner product is a linear map 
$$
L^\dagger :V  ⟶  V 
$$
defined by the formula 
$$
\left<   v , L^\dagger u \right> = \left<  L v , u \right>  
\quad \forall \quad 
u,v ∈ V
$$

\begin{exercise}
Let $V = 𝐑^n$ and take the inner product to be the Euclidean one. Let $L$ be a matrix and find the entries of $L^\dagger$ in terms of those of $L$.
\end{exercise}

Since $L^\dagger$ and $L$ are both operators on $V$, we can compare then. 
An operator is called self-adjoint if $L^\dagger = L$. 

\begin{exercise}
Let $v_λ$ and $v_μ$ be eigenvectors of $L$ and suppose $L$ is self-adjoint. 
Compare the “matrix elements” $\left< v_μ, L  v_λ\right> $ and $\left< v_μ, L^\dagger v_λ\right> $ to conclude that
$v_λ$ is orthogonal to $v_μ$ (with respect to the inner product) if $μ ≠ λ$. 
%$$
%0= \left< v_μ, (L-L^\dagger)  v_λ\right> 
%= (λ - μ) \left< v_μ,  v_λ\right> 
%$$
\end{exercise}

Consider the special case in which $\mathrm dim\ V = 2$ with the Euclidean inner product, and let $v₁ , v₂ ∈ V$ be  linearly-independent vectors. Compute the Euclidean length $|v₁|$, and define the unit vector 
$$
e₁ = { v₁ \over |v₁| }
$$
We will now make a new vector $ṽ₂$ defined by 
$$
ṽ₂ := v₂ - \left< v₂ , e₁ \right>  e₁
$$
It is orthogonal to $e₁$ since $e₁$ is a unit vector. Compute the length $|ṽ₂|$ and define the unit vector
$$
e₂ = { ṽ₂ \over |ṽ₂| }
$$
By construction 
$$
\left< e_i , e_j \right> = δ_{ij}
$$
so $V=\mathrm{Span}\{ e_i\}_{i=1}²$ is spanned by these “ortho(gonal)-normal(ized)” vectors. 


\begin{exercise}[\href{https://en.wikipedia.org/wiki/Gram–Schmidt_process}{Gram-Schmidt}]
Use mathematical induction to extend this result to a procedure applicable to any finite-dimensional inner-product space. This algorithm is called the Gram-Schmidt orthonormalization process. 
\end{exercise}


Together with the result on orthogonal eigenvectors for self-adjoint operators, we conclude that if we have a vector space equipped with such an operator, we can construct an orthonormal basis of eigenvectors for the vector space.


§{Function Spaces}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


¶{Orthogonal Polynomials}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Let $S⊂ 𝐑 $ be a subset of the real line and define the space $L²(S)$ of square-integrable functions on $S$:
$$
L²(S) = \left\{ 
		\ f : S ⟶ 𝐑 \ \Bigg| \int_S  |f(x)|² dx < ∞ 
	\right\}
$$
Note that $L²(S)$ is an inner-product space over $𝐑$ with 
\eq{\label{innerprod}
\left< g, f\right> = \int_S g(x) f(x) dx 
}

Consider $S=[-1,1]$. Note that any continuous function defined on such a set is automatically (square) integrable. 
\begin{exercise} 
Consider the monomials $x^n$ defined on $S$. Starting with $p₀=1$, show that $p₁ = x$ is orthogonal to it. Find $p₂
$ so that it is a quadratic polynomial orthogonal to both $p₀$ and $p₁$. Find $p₃$ so that it is a cubic polynomial orthogonal to $p_i$ with $i=0,1$, and 2.  
\end{exercise}
%\al{
%p₀ &= 1\cr
%p₁ &= x \cr
%p₂ &= 3x² -1\cr
%p₃ & = 5x³ - 3x
%}

\begin{exercise} 
Consider the operator 
\eq{\label{LegendreLaplacian}
L 
%= \frac{d}{dx} \left( (x² -1) \frac{d}{dx}\right)
= (x² -1) \frac{d²}{dx²} + 2x \frac{d}{dx} 
}
on $L²(S)$. 
Show that it is self-adjoint with respect to the inner product \eqref{innerprod}.\\
Hint: Show that any operator of the form $\frac{d}{dx} \left( p(x) \frac{d}{dx}\right)$ where $p(x)$ is a polynomial is “manifestly” self-adjoint, provided $p$ vanishes on the boundary of $S$. 
Then find $p(x)$ for the operator above. 
\end{exercise}

By convention, the Legendre polynomials $P_n$ are normalized by 
$$
\left< P_m , P_n\right>:= \int_{-1}¹ P_m(x) P_n(x)  dx = {2 \over 2n +1} δ_{mn}
$$
The Rodriguez formula gives them as
$$
P_n(x) = \frac1{2^nn!} \frac{d^n}{dx^n} (x² - 1)^n 
$$
These polynomials are the eigenvectors of the operator $L$ of \eqref{LegendreLaplacian}. 
\begin{exercise}
Check this on the first few Legendre polynomials $p_i$ for $i=0, ... , 3$. What are the eigenvalues? 
\end{exercise} 
%\al{
%λ₀ &= 0\cr
%λ₁ &= 2 \cr
%λ₂ &= 6\cr
%λ₃ & =  12
%}


\paragraph{Laplace equation}
The Legendre polynomials are famous because they are instrumental to the solution of the Laplace equation $Δ 𝜑  =0$ {\em in three dimensions}. In spherical coordinates, the three-dimensional Laplacian operator is 
$$
Δ = \frac {1}{r²} \frac{∂}{∂r} \left( r² \frac{∂}{∂r}\ ⋅ \right) 
	+\frac{1}{r² \sin θ } \frac{∂}{∂θ} \left( \sin θ  \frac{∂}{∂θ}\ ⋅ \right) 
	+\frac{1}{r² \sin θ } \frac{∂²}{∂φ²} 
$$
To solve a problem with spherical symmetry, the {\it Ansatz} is made that the solution is separable: $𝜑(r, θ, φ) = R(r) Θ(θ) 𝛷(φ)$.
This allows the separation of the Laplace equation into three parts. It can be shown that the $𝛷=$ constant and that the rest of the equation separates into the sum of two parts with one depending only on $r$ and the other only on $θ$. This means that the two must be equal to the same constant. The $θ$-dependent part is given by
$$
\frac{1}{\sin θ } \frac{d}{dθ} \left( \sin θ  \frac{d Θ}{dθ} \right)  + c Θ = 0. 
$$
(The separation constant is denoted by $c$.)

\begin{exercise} 
Make the change of variables $x = \cos θ$ to show that this is an eigenvalue equation for the Legendre operator \eqref{LegendreLaplacian}.
\end{exercise}

\paragraph{An aside on Taylor series} 
The completeness of eigenfunctions of a self-adjoint operator on an infinite-dimensional vector space is striking, but it should not be entirely shocking to a student of calculus. Recall that a real function $f$ of a real variable $x$ is analytic on a domain $D⊂ 𝐑$ if there is a convergent series on $D$ that agrees with the function $f$ on $D$:
$$
f(x) = \sum_{n=0}^∞ a_n x^n \quad \quad \textrm{on}\quad D.
$$
\begin{exercise} 
Find an expression for the coefficients $\{a_n\}_{n=0}^∞$ in terms of $f$ and its derivatives. 
\end{exercise}
%$$
%a_n = \frac{1}{n!} \left. \frac{d^n f}{dx^n} \right|_{x=0} .
%$$


¶{(Anti-)Periodic Functions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


Let $S=(-π, π)$, and consider the operator
$$
L = \frac{d²}{dx²}
$$
on $L²(S)$. Then 
$$
\int g \frac{d²}{dx²} f dx 
= \left. g \frac{df}{dx} \right|_{x=-π}^π  - \int \frac{dg}{dx}\frac{df}{dx}  dx 
= \left. g \frac{df}{dx} \right|_{x=-π}^π  
	- \left. \frac{dg}{dx} f \right|_{x=-π}^π  
	+\int \left(\frac{d²}{dx²} g\right)  f dx 
$$
The boundary terms will vanish if we restrict to the subset of $L²(S)$ of functions that vanish at $x =± π$ or have vanishing first derivatives at those points. Let's call the former $𝒟 (S)$ and the latter $𝒩 (S)$.  

\begin{exercise} 
Find the eigenvectors and eigenvalues of the self-adjoint operator $L$ on $𝒟(S)$ and $𝒩(S)$. Show that they are orthogonal and normalize them.
\end{exercise}

If the eigenvectors $\{v_n(x)\}_{n∈ 𝐙}$, say of a self-adjoint operator span the vector space on which that operator acts, then in the context of $L²(S)$, that must mean that it is possible to decompose {\em any} function $f∈ L²(S)$ in a (possibly infinite) sum of these eigenvectors $f(x) = \sum_{n∈ 𝐙} c_n v_n(x)$. Alternatively, 
\eq{\label{FourierReal}
f(x) = \sum_{n=0}^∞ \left[ 
	a_n \sin(n x) + b_n \cos(n x)
	\right]
}
for some series of real coefficients $a_n, b_n ∈ 𝐑\ \forall n$.
This form is called the Fourier series representation of the function $f(x)$. It gives an entire function in terms of a series $\{a_n\}_{n=1}^∞$ and $\{b_n\}_{n=0}^∞$ of real numbers called the Fourier coefficients. 

\begin{exercise} 
Assuming this decomposition of the function and using the orthogonality of the eigen-functions you found above, calculate the $a$ and $b$ coefficients. 
\end{exercise}



¶{Complex Hilbert Spaces}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Complex numbers are crucial for all life, so we would like to extend everything in sight to the complex domain. For one thing, that means that the vector spaces above will now be over $𝐂$. It turns out that the nicest inner product, however, should be real, or “Hermitian” to be more precise: It is linear in the second slot, and anti-linear in the first. 
Explicitly, 
$$
\left< v, a u\right>  = a \left< v,  u\right>  \quad \textrm{and} \quad \left< av, u\right>  = ā \left< v,  u\right>  
$$

\begin{exercise} 
Let $V$ be an $n$-dimensional inner product space over $𝐂$ with Hermitian form $\left<\ ⋅\ ,\ ⋅\ \right>$. 
Let $L : V ⟶ V$ be a linear operator on $V$. Then $L$ has a matrix representation. If $L$ is self-adjoint with respect to the inner product, what is the equivalent condition on its matrix representation? 
\end{exercise}
A self-adjoint operator on a Hermitian inner-product space is also called a Hermitian operator. 
\begin{exercise} 
Show that the eigenvalues of a Hermitian operator are all real. Show that eigenvectors with distinct eigenvalues are orthogonal. 
\end{exercise}

There is a great amount of very awesome stuff to discuss about complex variables. Unfortunately, we will have to skip all of it here. One thing we will need however is Euler's formula:
$$
e^{ix} = \cos (x) + i \sin (x)
$$
This is important to us for the following reason:

Consider extending the real Fourier analysis above to complex values (𝐂 range but 𝐑 domain). Then $f : 𝐑 ⟶ 𝐂$ will have complex coefficients in the analogue of its Taylor expansion. Thus, the coefficients $a$ and $b$ in \eqref{FourierReal} are now allowed to be complex. 
Mediate for a second on the conclusion that the complex analogue of that result is that for a complex-valued function $f$
\eq{\label{ComplexFourierSeries}
f(x) = \sum_{n∈ 𝐙} c_n e^{inx} 
}

\begin{exercise} 
For $m, n ∈ 𝐙$, show that\footnote{The symbol in the middle of the equation above is the Kronecker $δ$; it is defined by the expression on the right. }
$$
\frac{1}{2π} \int_{-π}^π  e^{- i m x}\ e^{inx}\ dx =\ δ_{m,n}\ :=  \begin{cases} 
		0 & \textrm{if}\quad m≠ n\\
		1 & \textrm{if}\quad m = n
	\end{cases} 
$$	
Use this formula to derive an expression for the Fourier coefficients $\{c_n\}_{n=0}^∞$.
\end{exercise} 


¶{Continuous Groups and Harmonic analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The complex Fourier series \eqref{ComplexFourierSeries} can be expressed as an integral over a group. In this case, the group $G=𝐙$ is discrete and the integral reduces to a sum, but in general, the group could be continuous and even non-abelian. 
This is clear already from the formula for the Fourier modes 
$$
c_n = \int_{𝐑} f(x) e^{-i n x} dx
$$

...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%







\end{document}  